How to select predictive models for decision-making or causal inference?

### 2025-02-07 main_submission.tex
prepare main_submission latex and pdf

Should we know the two potential outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two potential outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following
equality, linking the target quantity to the observed data:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$. Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

\begin{table*}[bt!]
    \makebox[\textwidth]{
        \begin{threeparttable}[b]
            \caption{Review of causal risks
                ---
                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
                \cite{schuler_comparison_2018}.
                \label{tab:evaluation_metrics}}
            \centering
            \begin{tabular}{llr}
                \toprule
                Risk                                                                                           & Equation
                                                                                                               & Reference                                                                                                                                           \\
                \midrule
                $\tau\text{-risk}=\text{MSE}(\tau(X), \tau_f(X))$                                              & $\mathbb E_{X\sim
                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                               & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                $\mu\text{-risk} = \text{MSE}(Y, f(X))$                                                        & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                               & Def. \ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                (Y-f(X ; A))^2 \right]$                                                                        & Def.
                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
                \\
                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
                \\
                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
                \\
                \bottomrule
            \end{tabular}
        \end{threeparttable}
    }
\end{table*}

@@

\begin{proposition}[$R \text{-risk}$ as reweighted
        $\tau\text{-risk}$]\label{theory:prop:r_risk_rewrite} Given an outcome model
    $f$, its $R\text{-risk}$ appears as weighted version of its $\tau\text{-risk}$
    (Proof in \ref{apd:proofs:r_risk_rewrite}):
    \begin{align}
        R\text{-risk}^*(f) & = \int_{x} e(x)\big(1-e(x)\big)\big(\tau(x)-\tau_ {f}(x)\big)^{2} p(x) d x \nonumber \\
                           & \quad\; + \tilde{\sigma}_B^2(1) + \tilde{\sigma}_B^{2}(0)
    \end{align}
\end{proposition}

The $R \text{-risk}$ targets the oracle at the cost of an overlap re-weighting
and the addition of the reweighted Bayes residuals, which are independent of
$f$. In good overlap regions the weights $e(x) \big(1-e(x) \big)$ are close to
$\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
$\tau \text{-risk}$. For randomized control trials, this weight is constant
making the $R\text{-risk}$ particularly suited for exploring
heterogeneity (Appendix \ref{apd:theory:special_cases})

@@

We use simulated data, on which the ground-truth causal effect is known. Going
beyond prior empirical studies of causal model selection
\cite{schuler_comparison_2018,alaa_validating_2019}, we use many
generative processes, which is needed to reach general conclusions (Appendix \ref{apd:results:seed_effect}).

We generate the response functions using random bases extension, a common method
in biostatistics, \eg functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, we control the complexity of the non-linear outcome
models. We use random approximation of Radial Basis Function (RBF) kernels
\cite{rahimi_random_2008} to generate the outcome and treatment functions. RBF
use the same process as polynomial splines but replace polynomial by Gaussian
kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
input space. This avoids unrealistic divergences of the functions at the
ends of the feature space. We generate 1\,000 datasets based on these functions,
with random overlap parameters. Example shown in Figure
\ref{fig:simulation_examples} and details in \ref{apd:experiments:generation}.

@@

We also use three semi-simulated
data adding a known synthetic causal effect to real --non synthetic-- healthcare
covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
developmental disorders containing 4,802 indivduals and 55 features. We used 770
dataset instances: 10 random seeds for each of the 77 simulated settings for the
treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
data on twin births and mortality rates \cite{almond_costs_2005}. There are
$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
different treatment allocations. Appendix
\ref{apd:experiments:datasets} gives datasets details.

@@

Drawing from the TMLE literature that uses combination of flexible
machine learning methods \cite{schuler_targeted_2017}, we model the nuisances
$\check e$ (respectivley $\check m$) with a meta-learner: a stacked estimator of ridge
and boosting classifiers (respectively regressions) (hyperparameter selection in Appendix \ref{apd:experiments:nuisances_hp}).

@@

%\idea{Overlap is important but not easily measurable}
Good overlap between treated and control population is crucial for causal
inference (Assumption \ref{assumption:overlap}). We introduce the
Normalized Total Variation (NTV), a divergence based on the propensity score
summarizing the overlap between both populations (Appendix
\ref{apd:motivation_ntv}).

\begin{figure}[!b]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_1_r_risk_domination_mrisks_by_ds_feasible.pdf}
    \end{minipage}
    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
        \ref{apd:table:relative_kendalls_all_datasets} gives median and
        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
\end{figure}

@@

Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the
different feasible causal metrics. We measure this agreement with relative
Kendall tau $\kappa$ (eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. To
remove the variance across datasets (some datasets lead to easier model
selection than others), we report values for one metric relative to the mean of
all metrics for a given dataset instance: $\text{Relative} \,
    \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
    mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$. Given the importance of
overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
separate strong and weak overlap.

@@

Beyond rankings, the differences in terms of absolute
ability to select the best model are large: The R-risk selects a model
with a $\tau\text{-risk}$ only 1\% higher
than the best
possible candidate for strong overlap on Caussim, but selecting with
the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
--as in some causal-inference methods \cite{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
(Figure
\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
datasets, the $R\text{-risk}$ consistently decreases the
risk compared to the $\mu\text{-risk}$: from
0.1\% to 1\% on ACIC2016,  1\% from to 20\% on ACIC2018,
and 0.05\% from to 1\% on Twins.

@@

Model selection for causal inference becomes more and more difficult with
increasingly different treated and control populations (Figure
\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
coefficient correlation with $\tau\text{-risk}$ drops from 0.9
(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
(\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}).

\begin{figure}[!h]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_2_overlap_influence_by_bin_k_by_ds_r_risk_2cols.pdf}
    \end{minipage}
    \caption{\textbf{Model selection is harder for low population
            overlap}:
        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. Supplementary materials presents results for all
        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
        Kendall's and continuous overlap values in Figure
            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
\end{figure}

@@

    \caption{\textbf{Nuisances can be estimated on the same data as outcome
            models}: Results for the R-risk are similar between the
        \textcolor{MidnightBlue}{shared
            nuisances/candidate set} and
        the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
        \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label{fig:procedures_comparison}
\end{figure}

@@

Stacked nuisances estimators (boosting and linear) lead to feasible
metrics with close performances to the oracles ones: the
corresponding estimators recover well-enough the true nuisances.
One may wonder if simpler models for the nuisance could be useful,
in particular in data-poor settings or when the true models are linear.
Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
selection estimating nuisances with stacked estimators or linear model.
It comprises the Twins data, where the true propensity model is linear,
and a downsampled version of this data, to study a situation favorable to
linear models. In these settings,
stacked and linear estimations of the nuisances performs equivalently.
Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
confirms that using adaptive models --as built by
stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.

@@

    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
                models} are good overall estimators of the nuisances}:
        Results are shown only for the
        R-risk; Figure \ref{apd:fig:nuisances_comparison}
        details every metrics. For Twins, where the true propensity
        model is linear, \textcolor{DarkOrchid}{stacked} and
        \textcolor{ForestGreen}{linear}
        estimations of the nuisances performs equivalently, even for a downsampled version
        (N=4,794). }\label{fig:all_datasets_nuisances_comparison}
\end{figure}

@@

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about treatment effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction. Rather, selecting the best outcome
model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
to more valid causal estimates on average. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \eg in machine
learning: it involves fitting nuisance models necessary for model
evaluation.
Our results show that these can be learned on the same set of data as the
outcome models evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. To select propensity score models, we used the Brier score,
minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

\appendix
\section{Additional files}

%\part*{Supplementary materials:\\How to select predictive models for decision-making or causal inference? }
%\setcounter{secnumdepth}{1}
%\renewcommand{\thesubsection}{\Alph{subsection}}
%\setcounter{secnumdepth}{3}

%\part{Supplementary materials}

%The following supporting information is available as part of the online article:

@@

Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
different models used in g-computation estimators on the 76 configurations of
the ACIC 2016 dataset. Outcome models are fitted on half of the data and
inference is done on the other half --ie. train/test with a split ratio of 0.5.
For each configuration, and each model, this train test split was repeated ten
times, yielding non parametric variance estimates
\cite{bouthillier_accounting_2021}. Figure \ref{fig:acic_2016_ate_heterogeneity}
shows large variations obtained across different outcome estimators on
semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models such as
random forests are doing well in most settings except when treated and untreated
populations differ noticeably, in which case a linear model (ridge) is to be
preferred. However random forests with different hyper-parameters (max depth= 2)
yield poor estimates. A simple rule of thumb such as preferring flexible models
does not work in general; model selection is needed.

%\idea{It is crucial to have tools for model selection.}

\begin{figure}[!b]
    \centering
    \includegraphics[width=0.95\linewidth]{acic_2016_ate_hte.pdf}%
    \caption{\textbf{Different outcome models lead to different
            estimation errors on the Average Treatment Effects},
        on 77 classic simulations with known true causal effect
        \cite{dorie_automated_2019}. The different models are ridge regression
        and random forests with different hyper-parameters
        (details
        \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
        plotted as a function of increasing difference between treated and
        untreated population --see
        \autoref{subsec:measuring_overlap}.
        There is no systematic best performer; data-driven model
        selection is important.
        \label{fig:acic_2016_ate_heterogeneity}%
    }
\end{figure}

Outcome models are implemented with
\href{https://scikit-learn.org/stable/}{scikit-learn}
\cite{pedregosa_scikitlearn_2011} and the following hyper-parameters:

\begin{table}[h!]
    \centering
    \begin{tabular}{llll}
        \toprule
        Outcome Model                                  & Hyper-parameters grid
        \\
        \midrule
        Random Forests                                 & Max depth: [2,
        10]                                                                    \\

        Ridge regression without treatment interaction & Ridge regularization:
        [0.1]                                                                  \\

        Ridge regression with treatment interaction    & Ridge regularization:
        [0.1]                                                                  \\
        \bottomrule
    \end{tabular}
    \caption{Hyper-parameters grid used for ACIC 2016 ATE variability}
    \label{apd:toy_example:acic_2016_ate_variability:table}
\end{table}

@@

A natural way to select a predictive model for causal inference would be
an error measure between a causal quantity such as the CATE and models' estimate. But such error is
not a ``feasible'' risk: it cannot be computed solely from observed data
and requires oracle knowledge.

% XXX: I think that I should make the following paragraph shorter

\paragraph{Simulation studies of causal model selection}

Using eight simulations setups from \cite{powers_methods_2018}, where
the oracle CATE is known, \cite{schuler_comparison_2018} compare four
causal risks, concluding that for CATE estimation the best
model-selection risk is the so-called $R\text{-risk}$
\cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
empirical results are clear for randomized treatment allocation but less
convincing for observational settings where both simple MSE --$\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
appear to perform better than $R\text{-risk}$ on half of the simulations.
Another work \cite{alaa_validating_2019} studied empirically both MSE and
reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
prior empirical work by studying a wider variety of data generative
processes and varying the influence of overlap, an important parameter of
the data generation process which makes a given causal metric appropriate
\cite{damour_overlap_2020}. We also study how to best adapt
cross-validation procedures to causal metrics which themselves come with
models to estimate.

\paragraph{Theoretical studies of causal model selection}

Several theoretical works have proposed causal model selection procedures
that are \emph{consistent}: select the best model in a family given
asymptotically large data. These work rely on introducing a
CATE estimator in the testing procedure: matching
\cite{rolling_model_2014}, an IPW estimate
\cite{gutierrez_causal_2016}, a doubly robust estimator
\cite{saito_counterfactual_2020}, or debiasing the error with influence
functions \cite{alaa_validating_2019}. However, for theoretical
guarantees to hold, the test-set correction needs to converge to the
oracle: it needs to be flexible enough --well-posed-- and asymptotic
data. From a practical perspective, meeting such requirements
implies having a good CATE estimate, thus having solved
the original problem of causal model selection.

\paragraph{Statistical guarantees on causal estimation procedures}

Much work in causal inference has focused on procedures that
guarantee asymptotically consistent estimators, such as Targeted
Machine Learning
Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} or
Double Machine Learning \cite{chernozhukov_double_2018}. Here also, theories require asymptotic regimes and
models to be \textit{well-specified}.

By contrast, without assuming that estimators are well specified, there exists an upper bound
on the oracle error to the CATE ($\tau\text{-risk}$) that involves the error on
the outcome and the similarity of the distributions of treated and control
patients \cite{johansson2022generalization}. However, they use this upper bound for model optimization,
and do not give insights on model selection. In addition, for hyperparameter
selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
counterfactual nearest neighbors, which has been shown ineffective
\cite{schuler_comparison_2018}. An interesting direction is taken in \cite{foster2023orthogonal} where the authors derive convergence rates for orthogonal losses such as the R-loss, or the DR-Loss without assuming well-specification of the model for target parameter.

>>>>

Should we know the two potential outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two potential outcomes can be estimated from observed data:
the consistency and unconfoundedness
\ref{apd:causal_assumptions} assumptions imply the following
equality, linking the target quantity to the observed data:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$. Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

\begin{table*}[bt!]
    \makebox[\textwidth]{
        \begin{threeparttable}[b]
            \caption{Review of causal risks
                ---
                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
                \cite{schuler_comparison_2018}.
                \label{tab:evaluation_metrics}}
            \centering
            \begin{tabular}{llr}
                \toprule
                Risk                                                                                           & Equation
                                                                                                               & Reference                                                                                                                                           \\
                \midrule
                $\tau\text{-risk}=\text{MSE}(\tau(X), \tau_f(X))$                                              & $\mathbb E_{X\sim
                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                               & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                $\mu\text{-risk} = \text{MSE}(Y, f(X))$                                                        & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                               & \cite{schuler_comparison_2018}                                                                                                                      \\
                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                (Y-f(X ; A))^2 \right]$                                                                        & \cite{vanderlaan_unified_2003}                                                                                                                      \\
                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                \cite{wager_estimation_2018}
                \\
                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                \cite{nie_quasioracle_2017}
                \\
                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                \cite{nie_quasioracle_2017}
                \\
                \bottomrule
            \end{tabular}
        \end{threeparttable}
    }
\end{table*}

@@

\begin{proposition}[$R \text{-risk}$ as reweighted
        $\tau\text{-risk}$]\label{theory:prop:r_risk_rewrite} Given an outcome model
    $f$, its $R\text{-risk}$ appears as weighted version of its $\tau\text{-risk}$
    (Proof in Appendix \ref{apd:proofs}):
    \begin{align}
        R\text{-risk}^*(f) & = \int_{x} e(x)\big(1-e(x)\big)\big(\tau(x)-\tau_ {f}(x)\big)^{2} p(x) d x \nonumber \\
                           & \quad\; + \tilde{\sigma}_B^2(1) + \tilde{\sigma}_B^{2}(0)
    \end{align}
\end{proposition}

The $R \text{-risk}$ targets the oracle at the cost of an overlap re-weighting
and the addition of the reweighted Bayes residuals, which are independent of
$f$. In good overlap regions the weights $e(x) \big(1-e(x) \big)$ are close to
$\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
$\tau \text{-risk}$. For randomized control trials, this weight is constant
making the $R\text{-risk}$ particularly suited for exploring
heterogeneity (Appendix \ref{apd:proofs})

@@

We use simulated data, on which the ground-truth causal effect is known. Going
beyond prior empirical studies of causal model selection
\cite{schuler_comparison_2018,alaa_validating_2019}, we use many
generative processes, which is needed to reach general conclusions (Appendix \ref{apd:experiments:additional_results}).

We generate the response functions using random bases extension, a common method
in biostatistics, \eg functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, we control the complexity of the non-linear outcome
models. We use random approximation of Radial Basis Function (RBF) kernels
\cite{rahimi_random_2008} to generate the outcome and treatment functions. RBF
use the same process as polynomial splines but replace polynomial by Gaussian
kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
input space. This avoids unrealistic divergences of the functions at the
ends of the feature space. We generate 1\,000 datasets based on these functions,
with random overlap parameters. Example shown in Appendix
\ref{apd:experiments:additional_results}, Fig  13. and details in Appendix \ref{apd:experiments:additional_results}.

@@

We also use three semi-simulated
data adding a known synthetic causal effect to real --non synthetic-- healthcare
covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
developmental disorders containing 4,802 indivduals and 55 features. We used 770
dataset instances: 10 random seeds for each of the 77 simulated settings for the
treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
data on twin births and mortality rates \cite{almond_costs_2005}. There are
$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
different treatment allocations. Appendix
\ref{apd:experiments:additional_results} gives datasets details.

@@

Drawing from the TMLE literature that uses combination of flexible
machine learning methods \cite{schuler_targeted_2017}, we model the nuisances
$\check e$ (respectivley $\check m$) with a meta-learner: a stacked estimator of ridge
and boosting classifiers (respectively regressions) (hyperparameter selection in Appendix \ref{apd:experiments:additional_results}).

@@

%\idea{Overlap is important but not easily measurable}
Good overlap between treated and control population is crucial for causal
inference (Assumption \ref{apd:causal_assumptions}). We introduce the
Normalized Total Variation (NTV), a divergence based on the propensity score
summarizing the overlap between both populations (Appendix
\ref{apd:motivation_ntv}).

\begin{figure}[!b]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_1_r_risk_domination_mrisks_by_ds_feasible.pdf}
    \end{minipage}
    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
        Normalized Total Variation (Appendix \ref{apd:motivation_ntv}). Appendix \ref{apd:experiments:additional_results} presents the same results
        by adding semi-oracle risks in Figure 14, measured with
        absolute Kendall's in Figure 15 and with $\tau\mathrm{-risk}$
        gains in Figure 16. Table
        4 gives median and
        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
\end{figure}

@@

Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the
different feasible causal metrics. We measure this agreement with relative
Kendall tau $\kappa$ (see Appendix \ref{apd:experiments:additional_results}, eq. 20) \cite{kendall_new_1938}. To
remove the variance across datasets (some datasets lead to easier model
selection than others), we report values for one metric relative to the mean of
all metrics for a given dataset instance: $\text{Relative} \,
    \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
    mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$. Given the importance of
overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
separate strong and weak overlap.

@@

Beyond rankings, the differences in terms of absolute
ability to select the best model are large: The R-risk selects a model
with a $\tau\text{-risk}$ only 1\% higher
than the best
possible candidate for strong overlap on Caussim, but selecting with
the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
--as in some causal-inference methods \cite{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
(Appendix \ref{apd:experiments:additional_results}, Figure 16). Across
datasets, the $R\text{-risk}$ consistently decreases the
risk compared to the $\mu\text{-risk}$: from
0.1\% to 1\% on ACIC2016,  1\% from to 20\% on ACIC2018,
and 0.05\% from to 1\% on Twins.

@@

Model selection for causal inference becomes more and more difficult with
increasingly different treated and control populations (Figure
\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
coefficient correlation with $\tau\text{-risk}$ drops from 0.9
(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
(Appendix \ref{apd:experiments:additional_results}, Figure 15).

\begin{figure}[!h]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_2_overlap_influence_by_bin_k_by_ds_r_risk_2cols.pdf}
    \end{minipage}
    \caption{\textbf{Model selection is harder for low population
            overlap}:
        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
        are the tertiles of the overlap measured with NTV (Appendix \ref{apd:motivation_ntv}, eq. 17). Appendix \ref{apd:experiments:additional_results} presents results for all
        metrics in Figure 18 in absolute
        Kendall's and continuous overlap values in Figure
            {15}.}\label{fig:all_datasets_overlap_effect_r_risk}
\end{figure}

@@

    \caption{\textbf{Nuisances can be estimated on the same data as outcome
            models}: Results for the R-risk are similar between the
        \textcolor{MidnightBlue}{shared
            nuisances/candidate set} and
        the \textcolor{RedOrange}{separated nuisances set} procedures. Appendix \ref{apd:experiments:additional_results} Figure
        17 details results for all metrics.}\label{fig:procedures_comparison}
\end{figure}

@@

Stacked nuisances estimators (boosting and linear) lead to feasible
metrics with close performances to the oracles ones: the
corresponding estimators recover well-enough the true nuisances.
One may wonder if simpler models for the nuisance could be useful,
in particular in data-poor settings or when the true models are linear.
Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
selection estimating nuisances with stacked estimators or linear model.
It comprises the Twins data, where the true propensity model is linear,
and a downsampled version of this data, to study a situation favorable to
linear models. In these settings,
stacked and linear estimations of the nuisances performs equivalently.
Detailed analysis (Appendix \ref{apd:experiments:additional_results} Figure 20)
confirms that using adaptive models --as built by
stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.

@@

    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
                models} are good overall estimators of the nuisances}:
        Results are shown only for the
        R-risk; \ref{apd:experiments:additional_results} Figure 19
        details every metrics. For Twins, where the true propensity
        model is linear, \textcolor{DarkOrchid}{stacked} and
        \textcolor{ForestGreen}{linear}
        estimations of the nuisances performs equivalently, even for a downsampled version
        (N=4,794). }\label{fig:all_datasets_nuisances_comparison}
\end{figure}

@@

\paragraph{Nuisance models: more gain than pain}
%
Predictive models are increasingly used to reason about treatment effects, for
instance in precision medicine to drive individualized decision. Our results
highlight that they should be selected, validated, and tuned using different
procedures and error measures than those classically used to assess prediction. Rather, selecting the best outcome
model according to the $R\text{-risk}$ (Appendix \ref{def:feasible_risks} eq.\, Definition 5) leads
to more valid causal estimates on average. Estimating the $R\text{-risk}$ requires a more
complex procedure than standard cross-validation used \eg in machine
learning: it involves fitting nuisance models necessary for model
evaluation.
Our results show that these can be learned on the same set of data as the
outcome models evaluated. The nuisance models must be well estimated (Figure
\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
nuisance models a flexible stacking-based family of estimator suffices for good
model selection. To select propensity score models, we used the Brier score,
minimized by the true individual probability. An easy mistake is to
use calibration errors popular in machine learning
\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
as these select not for the individual posterior probability but for an
aggregate error rate \cite{perez2022beyond}.

@@

\appendix
\section{List of additional files}

@@

@@
